Simulating Ultra High-Dimensional Multivariate Data

### 2021-02-28 notebooks/bigsimr-julia-getting-started.Rmd
re-add notebooks so Grant can check out new branch without losing files

```{r setup}
## devtools::install_github("SchisslerGroup/bigsimr", ref="julia")
library(bigsimr)
Sys.setenv(JULIA_NUM_THREADS=parallel::detectCores())
## Either let JuliaCall find your installation of Julia or set manually
bs <- bigsimr_setup(JULIA_HOME = "/usr/local/bin/")
dist <- distributions_setup() # Julia distributions
library(tidyverse)
library(patchwork)
```

>>>>

```{r setup}
## devtools::install_github("SchisslerGroup/bigsimr", ref="julia")
library(bigsimr)
Sys.setenv(JULIA_NUM_THREADS=parallel::detectCores())
## Either let JuliaCall find your installation of Julia or set manually
## bs <- bigsimr_setup(JULIA_HOME = "/usr/local/bin/")
bs <- bigsimr_setup()
dist <- distributions_setup() # Julia distributions
library(tidyverse)
library(patchwork)
```

### 2021-02-28 notebooks/bivariateGamma.rmd
re-add notebooks so Grant can check out new branch without losing files

```{r LoadLib, echo = TRUE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
library(ggplot2)
library(tidyverse)
library(knitr)
library(dplyr)
knitr::opts_chunk$set(cache=TRUE, autodep=TRUE, cache.comments=TRUE)
set.seed(06032020)
library(bigsimr)
CORES <- parallel::detectCores() - 1
## CORES <- 1
## devtools::install_github( "superdesolator/NORTARA" )
library(nortaRA)
```

@@

```{r, globalParams}
## bivariate gamma
d <- 2
shape1 <- 5
shape2 <- 5
margins <- list(
    list("gamma", shape = shape1),
    list("gamma", shape = shape2)
)

@@

## storing the results
boundsDat <- NULL
simRes <- NULL
```

@@

```{r, boundsKendall, eval=FALSE}
type <- 'kendall'
## Using GSC method of Demirtas2011
(tmpBounds <- computeCorBounds( params = margins, cores = CORES, type = type, reps = N ))
tmpBoundsDat <- data.frame( shape1 = shape1, shape2 = shape2, type = type, upper = tmpBounds$upper[ 1, 2 ], lower = tmpBounds$lower[ 1, 2 ] )
boundsDat <- rbind( boundsDat, tmpBoundsDat )
```

@@

```{r bigsimrPearson, eval = TRUE, results = 'hide'}
type <- 'pearson'
## now show that our simulation methods works
tmpLower <- boundsDat$lower[ boundsDat$type == type ]
tmpUpper <- boundsDat$upper[ boundsDat$type == type ]
rVec <- seq( tmpLower, tmpUpper, length.out = gridSteps)  ## loose approximation for now

## loop for simplicity see if fast enough
for (r in rVec) {
    print( paste0( 'Working on true rho of ', r ) )
    rho <- matrix(r, nrow = d, ncol = d)
    diag(rho) <- 1.0
    simData  <- as.data.frame( rvec(n = N, rho = rho, params = margins, type = type, cores = CORES, adjustForDiscrete = FALSE) )
    tmpSimData <- data.frame( shape1 = shape1, shape2 = shape2, trueRho = r, estRho = cor(simData, method = type)[ 1, 2], method = 'bigsimr' , type = type)
    simRes <- rbind( simRes, tmpSimData )
}
```

@@

```{r bigsimrSpearman, eval = TRUE, results = 'hide'}
type <- 'spearman'
## now show that our simulation methods works
tmpLower <- boundsDat$lower[ boundsDat$type == type ]
tmpUpper <- boundsDat$upper[ boundsDat$type == type ]
rVec <- seq( tmpLower, tmpUpper, length.out = gridSteps)  ## loose approximation for now

## loop for simplicity see if fast enough
for (r in rVec) {
    print( paste0( 'Working on true rho of ', r ) )
    rho <- matrix(r, nrow = d, ncol = d)
    diag(rho) <- 1.0
    simData  <- as.data.frame( rvec(n = N, rho = rho, params = margins, type = type, cores = CORES, adjustForDiscrete = FALSE) )
    tmpSimData <- data.frame( shape1 = shape1, shape2 = shape2, trueRho = r, estRho = cor(simData, method = type)[ 1, 2], method = 'bigsimr' , type = type)
    simRes <- rbind( simRes, tmpSimData )
}
```

@@

```{r bigsimrKendall, eval = FALSE}
type <- 'kendall'
## now show that our simulation methods works
tmpLower <- boundsDat$lower[ boundsDat$type == type ]
tmpUpper <- boundsDat$upper[ boundsDat$type == type ]
rVec <- seq( tmpLower, tmpUpper, length.out = gridSteps )

## loop for simplicity see if fast enough
for (r in rVec) {
    print( paste0( 'Working on true rho of ', r ) )
    rho <- matrix(r, nrow = d, ncol = d)
    diag(rho) <- 1.0
    simData  <- as.data.frame( rvec(n = N, rho = rho, params = margins, type = type, cores = CORES, adjustForDiscrete = FALSE) )
    tmpSimData <- data.frame( shape1 = shape1, shape2 = shape2, trueRho = r, estRho = cor(simData, method = type)[ 1, 2], method = 'bigsimr' , type = type)
    simRes <- rbind( simRes, tmpSimData )
}
```

## `NORTARA` simulations

```{r nortaraSetup}
## slightly different input format than bigsimr
invcdfnames <- c("qgamma","qgamma")
paramslists <- list(
    m1 = list(shape = shape1 ),
    m2 = list(shape = shape2 )
)
```

@@

```{r nortaraPearson, eval = TRUE, warning=FALSE, results='hide'}
type <- 'pearson'
## now show that our simulation methods works
tmpLower <- boundsDat$lower[ boundsDat$type == type ]
tmpUpper <- boundsDat$upper[ boundsDat$type == type ]
esp <- 1e-3
rVec <- seq( tmpLower+esp, tmpUpper-esp, length.out = gridSteps ) 
## drop N for this one
tmpN <- 1e4
## loop for simplicity see if fast enough
for (r in rVec) {
    print( paste0( 'Working on true rho of ', r ) )
    rho <- matrix(r, nrow = d, ncol = d)
    diag(rho) <- 1.0
    simData  <- nortaRA::gennortaRA(n = tmpN, cor_matrix = rho, invcdfnames, paramslists)
    tmpSimData <- data.frame( shape1 = shape1, shape2 = shape2, trueRho = r, estRho = cor(simData, method = type)[ 1, 2], method = 'nortara' , type = type)
    simRes <- rbind( simRes, tmpSimData )
}
```

@@

```{r empVspecPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.path='fig/plot-', out.width= '75%', fig.align='center', fig.ext='pdf', fig.cap = 'Plot of the specified (true) versus empirical (computed) correlations for bivariate lognormal data with $p_1 = p_2 = 0.2$ generated by the standard approach, bigsimr, and T-Poisson. Empirical correlations are the estimated correlations from the simulated samples.'}
p0 <- ggplot(data = simRes, aes(x = trueRho, y = estRho, color = factor(method)) )
p0 + geom_point() + facet_grid(. ~ type ) + geom_abline(slope = 1, intercept = 0 )
```

>>>>

```{r LoadLib, echo = TRUE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
library(ggplot2)
library(tidyverse)
library(knitr)
library(dplyr)
knitr::opts_chunk$set(cache=TRUE, autodep=TRUE, cache.comments=TRUE)
set.seed(06032020)
## devtools::install_github( "adknudson/bigsimr", ref = 'develop')
library(bigsimr)
## devtools::install_github( "superdesolator/NORTARA" )
library(nortaRA)
## to parallize
library(parallel)
CORES <- parallel::detectCores() - 1
```

@@

```{r, globalParams, results = "hide"}
## bivariate gamma
d <- 2
shape1 <- 5
shape2 <- 5
margins <- list(
    list("gamma", shape = shape1),
    list("gamma", shape = shape2)
)

@@

## storing the results
boundsDat <- NULL
simRes <- NULL

## parallelize outside of bigsimr
cl <- makeCluster(CORES)
clusterEvalQ(cl, library("bigsimr"))
```

```{r runSim}
runSim <- function( r, N, margins, cores, simMethod, corType, invcdfnames = NULL) {
    d <- length(margins)
    print( paste0( 'Working on true rho of ', r ) )
    rho <- matrix(r, nrow = d, ncol = d)
    diag(rho) <- 1.0
    if (simMethod == 'bigsimr') {
        simData  <- as.data.frame( rvec(n = N, rho = rho, params = margins, type = corType, cores = cores) )
    }
    if (simMethod == 'nortaRA') {
        simData <- nortaRA::gennortaRA(n = N, cor_matrix = rho, invcdfnames, margins)
    } 
    tmpSimData <- data.frame( shape1 = margins[[1]]$shape, shape2 = margins[[2]]$shape, trueRho = r, estRho = fastCor(simData, method = type)[ 1, 2], method = simMethod , type = corType)
    return(tmpSimData)
}
## runSim( r = 0.5, N = 1e2, margins = margins, cores = 1, simMethod = 'bigsimr', corType = 'pearson')
```

@@

```{r, boundsKendall, eval=TRUE}
type <- 'kendall'
## Using GSC method of Demirtas2011
(tmpBounds <- computeCorBounds( params = margins, cores = CORES, type = type, reps = N ))
tmpBoundsDat <- data.frame( shape1 = shape1, shape2 = shape2, type = type, upper = tmpBounds$upper[ 1, 2 ], lower = tmpBounds$lower[ 1, 2 ] )
boundsDat <- rbind( boundsDat, tmpBoundsDat )
```

@@

```{r bigsimrPearson, eval = TRUE, results = 'hide'}
simMethod <- 'bigsimr'
type <- 'pearson'

## now show that our simulation methods works
tmpLower <- boundsDat$lower[ boundsDat$type == type ]
tmpUpper <- boundsDat$upper[ boundsDat$type == type ]
## rVec <- c(-0.4, 0.1, 0.5)
## rVec <- seq( tmpLower, tmpUpper, length.out = gridSteps)
## loose on the boundary for now
epsilon <- 1e-2
rVec <- seq( tmpLower+epsilon, tmpUpper-epsilon, length.out = gridSteps )

clusterExport(cl, varlist = c( 'simMethod', 'type') )
tmpSimRes <- as.data.frame( do.call( "rbind", parSapply( cl = cl, rVec, FUN = runSim, N = N, margins = margins, cores = 1, simMethod = simMethod, corType = type, simplify = FALSE) ) )
simRes <- rbind( simRes, tmpSimRes )

@@

```{r bigsimrSpearman, eval = TRUE, results = 'hide'}
type <- 'spearman'
simMethod <- 'bigsimr'

## now show that our simulation methods works
tmpLower <- boundsDat$lower[ boundsDat$type == type ]
tmpUpper <- boundsDat$upper[ boundsDat$type == type ]
## rVec <- c(-0.4, 0.1, 0.5)
## rVec <- seq( tmpLower, tmpUpper, length.out = gridSteps)
## loose on the boundary for now
epsilon <- 1e-2
rVec <- seq( tmpLower+epsilon, tmpUpper-epsilon, length.out = gridSteps )

clusterExport(cl, varlist = c( 'simMethod', 'type') )
tmpSimRes <- as.data.frame( do.call( "rbind", parSapply( cl = cl, rVec, FUN = runSim, N = N, margins = margins, cores = 1, simMethod = simMethod, corType = type, simplify = FALSE) ) )
simRes <- rbind( simRes, tmpSimRes )

@@

```{r bigsimrKendall, eval = TRUE}
type <-'kendall'
simMethod <- 'bigsimr'

## now show that our simulation methods works
tmpLower <- boundsDat$lower[ boundsDat$type == type ]
tmpUpper <- boundsDat$upper[ boundsDat$type == type ]
## rVec <- c(-0.4, 0.1, 0.5)
## rVec <- seq( tmpLower, tmpUpper, length.out = gridSteps)
## loose on the boundary for now
epsilon <- 1e-2
rVec <- seq( tmpLower+epsilon, tmpUpper-epsilon, length.out = gridSteps )

clusterExport(cl, varlist = c( 'simMethod', 'type') )
tmpSimRes <- as.data.frame( do.call( "rbind", parSapply( cl = cl, rVec, FUN = runSim, N = N, margins = margins, cores = 1, simMethod = simMethod, corType = type, simplify = FALSE) ) )
simRes <- rbind( simRes, tmpSimRes )
```

## `nortaRA` simulations

```{r nortaRAsetup, eval = TRUE}
## slightly different input format than bigsimr
invcdfnames <- c("qgamma","qgamma")
paramslists <- list(
    m1 = list(shape = shape1 ),
    m2 = list(shape = shape2 )
)
```

@@

```{r nortaRApearson, eval = TRUE, warning=FALSE, results='hide'}
type <- 'pearson'
simMethod <- 'nortaRA'

## now show that our simulation methods works
tmpLower <- boundsDat$lower[ boundsDat$type == type ]
tmpUpper <- boundsDat$upper[ boundsDat$type == type ]
## rVec <- c(-0.4, 0.1, 0.5)
## rVec <- seq( tmpLower, tmpUpper, length.out = gridSteps)
## loose on the boundary for now
epsilon <- 1e-2
rVec <- seq( tmpLower+epsilon, tmpUpper-epsilon, length.out = gridSteps )

## drop N for this one
tmpN <- 1e4

## runSim( r = 0.4, N = tmpN, margins = paramslists, cores = 1, simMethod = simMethod, corType = type, invcdfnames = invcdfnames)
clusterExport(cl, varlist = c( 'simMethod', 'type') )
tmpSimRes <- as.data.frame( do.call( "rbind", parSapply( cl = cl, rVec, FUN = runSim, N = N, margins = paramslists, cores = 1, simMethod = simMethod, corType = type, invcdfnames = invcdfnames, simplify = FALSE) ) )
simRes <- rbind( simRes, tmpSimRes )
```

@@

```{r empVspecPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.path='fig/plot-', out.width= '75%', fig.align='center', fig.ext='pdf', fig.cap = 'Plot of the specified versus empirical correlations for standard bivariate gamma data with shape1 = shape2 = 5 generated by various approaches. Empirical correlations are the estimated correlations from the simulated samples.'}
p0 <- ggplot(data = simRes, aes(x = trueRho, y = estRho, color = factor(method)) )
p0 + geom_point() + facet_grid(. ~ type ) + geom_abline(slope = 1, intercept = 0 )
```

```{r cleanUP, echo = FALSE}
## When you're done, clean up the cluster
## stopCluster(cl)
```

### 2021-02-28 notebooks/cmstats2020_talk.rmd
re-add notebooks so Grant can check out new branch without losing files

- We propose a Gaussian-copula random vector algorithm; an augmented NORmal To Anything (NORTA) @Cario1997 method. 
- We provide a high-performance and GPU-accelerated implementation in the `R` package `bigsimr`, along with related tools.
- An article motivating, describing, and applying the methods; archived [here](???).

@@

| Simulation Reps ($B$) | Correlation Types | Identical-margin 2D distribution |
|-------------|:--------------:|----------------------:|
|$1000$ | Pearson ($\rho_P$) | ${ \bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho )$ |
|$10,000$ | Spearman ($\rho_S$) | ${ \bf Y} \sim MVG( shape = 10, rate = 1, \rho )$ |
| $100,000$| Kendall ($\tau$)| ${ \bf Y} \sim MVNB(p = `r prob`, r = 4,\rho)$ |

- For each of the unique 9 simulation configurations above, we estimate the correlation bounds and vary $\rho$ along a sequence of 100 points evenly placed within the bounds.

@@

```{r ch050-estRhoBRCA, echo=TRUE, eval=TRUE, cache=F}
## Estimate Spearman's correlation on the count data
corType <- 'spearman'
system.time( nb_Rho <- bigsimr::cor_fast( brca, method = corType ) )
```
## 4B - Estimate the marginal parameters

@@

$$
\hat{P}( {\bf Y} >= {\bf y_0 } ) = \sum_{b=1}^B I( {\bf Y^{(b) }} > {\bf y_0 } ) / B
$$

@@

```{r ch050-densityEvaluation, echo=TRUE, eval=TRUE, cache=F}
d <- ncol(sim_nbinom)
B <- nrow(sim_nbinom)
threshold <- rep( 1, d)
mean(apply( sim_nbinom,  1,
           function(X, y0=threshold) {
               all( X > y0) }
           ))
```

>>>>

- We propose a Gaussian-copula random vector algorithm; an augmented NORmal To Anything (NORTA) @Cario1997 method. 
- We provide a high-performance and GPU-accelerated implementation in the `R` package `bigsimr`, along with related tools.
- An article motivating, describing, and applying the methods; in development.

@@

| Simulation Reps ($B$) | Correlation Types | Identical-margin 2D distribution |
|-------------|:--------------:|----------------------:|
|$1000$ | Pearson ($\rho_P$) | ${ \bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho_i ), i=1,\ldots,100$ |
|$10,000$ | Spearman ($\rho_S$) | ${ \bf Y} \sim MVG( shape = 10, rate = 1, \rho_i ), i=1,\ldots,100$ |
| $100,000$| Kendall ($\tau$)| ${ \bf Y} \sim MVNB(p = `r prob`, r = 4,\rho_i), i=1,\ldots,100$ |

- For each of the unique 9 simulation configurations described above, we estimate the correlation bounds and vary the correlations along a sequence of 100 points evenly placed within the bounds, aiming to explore $\Omega$.  
- Specifically, we set correlations $\{ \rho_1 = ( \hat{l} + \epsilon), \rho_2 = (\hat{l} + \epsilon) + \delta, \ldots, \rho_{100} = (\hat{u} - \epsilon) \}$, with $\hat{l}$ and $\hat{u}$ being the estimated lower and upper bounds, respectively, and increment value $\delta$.  
- The adjustment factor, $\epsilon=0.01$, is introduced to handle numeric issues when the bound is specified exactly.

@@

```{r ch050-estRhoBRCA, echo=TRUE, eval=TRUE, cache=F}
## Estimate Spearman's correlation on the count data
corType <- 'spearman'
system.time( nb_Rho <- bigsimr::cor_fast( brca, method = corType ) )
```

@@

$$
\hat{P}( {\bf Y} \geq {\bf y_0 } ) = \frac{1}{B} \sum_{b=1}^B I( {\bf Y^{(b) }} \geq {\bf y_0 } )
$$

@@

```{r ch050-densityEvaluation, echo=TRUE, eval=TRUE, cache=F}
d <- ncol(sim_nbinom)
B <- nrow(sim_nbinom)
threshold <- rep( 1, d)
mean(apply( sim_nbinom,  1,
           function(X, y0=threshold) {
               all( X >= y0) }
           ))
```

### 2021-02-28 notebooks/esco2020_talk.rmd
re-add notebooks so Grant can check out new branch without losing files

---
title: "Simulating Ultra High-Dimensional Multivariate Data"
subtitle: "Using the `bigsimr` R package"
author: "A. Grant Schissler, Alexander Knudson, Tomasz J. Kozubowski, Anna K. Panorska"
institute: |
    "Department of Mathematics & Statistics" |
    "University of Nevada, Reno"
date: "8 Jun 2020 (updated: `r Sys.Date()`)"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
    incremental: true
    logo: ../man/figures/logo.png
bibliography: /Users/alfred/Dropbox/bib/bigsimr.bib
abstract: "In this era of Big Data, it is critical to realistically simulate data to conduct informative Monte Carlo studies. This is problematic when data are inherently multivariate while at the same time are (ultra-) high dimensional. This situation appears frequently in observational data found on online and in high-throughput biomedical experiments (e.g., RNA-sequencing). Due to the difficulty in simulating realistic correlated data points, researchers often resort to simulation designs that posit independence --- greatly diminishing the insight into the empirical operating characteristics of any proposed methodology.  Major challenges lie in the computational complexity involved in simulating these massive random vectors. We propose a fairly general, scalable procedure to simulate high-dimensional multivariate distributions with pre-specified marginal characteristics and dependency characteristics. As a motivating example, we use our methodology to study large-scale statistical inferential procedures applied to cancer-related RNA-sequencing data sets. The proposed algorithm is implemented as the bigsimr R package."
---

@@

- `bigsimr` is on `Github`: [https://github.com/adknudson/bigsimr](https://github.com/adknudson/bigsimr).
- Use `devtools` to install.
- `bigsimr` uses `JAX` python libraries for GPU acceleration.

>>>>

---
title: "Simulating Ultra High-Dimensional Multivariate Data"
subtitle: "Using the `bigsimr` R package"
author: "A. Grant Schissler, Alexander Knudson, Tomasz J. Kozubowski, Anna K. Panorska"
institute: |
    "Department of Mathematics & Statistics" |
    "University of Nevada, Reno"
date: "8 Jun 2020 (updated: `r Sys.Date()`)"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
    logo: ../man/figures/logo.png
bibliography: /Users/alfred/Dropbox/bib/bigsimr.bib
abstract: "In this era of Big Data, it is critical to realistically simulate data to conduct informative Monte Carlo studies. This is problematic when data are inherently multivariate while at the same time are (ultra-) high dimensional. This situation appears frequently in observational data found on online and in high-throughput biomedical experiments (e.g., RNA-sequencing). Due to the difficulty in simulating realistic correlated data points, researchers often resort to simulation designs that posit independence --- greatly diminishing the insight into the empirical operating characteristics of any proposed methodology.  Major challenges lie in the computational complexity involved in simulating these massive random vectors. We propose a fairly general, scalable procedure to simulate high-dimensional multivariate distributions with pre-specified marginal characteristics and dependency characteristics. As a motivating example, we use our methodology to study large-scale statistical inferential procedures applied to cancer-related RNA-sequencing data sets. The proposed algorithm is implemented as the bigsimr R package."
---

@@

- `bigsimr` is on `Github`: [https://github.com/adknudson/bigsimr](https://github.com/adknudson/bigsimr).
- Use `devtools` to install.
- `bigsimr` uses `JAX` python libraries for GPU acceleration and other high
  performance algorithms.
